# Univariate peaks-over-threshold analysis


This example illustrates some of the functions used in peaks-over-threshold analysis based on fitting a generalized Pareto distribution to threshold exceedances. We use the Venezuelian rainfall data, a time series of daily rainfall precipitations at Maiquetia airport in Venezuela, for the purpose of illustration.


```{r "thresholdstab"}
library(mev)
library(lubridate, quietly = TRUE, warn.conflicts = FALSE)

data("maiquetia")
day <- seq.Date(from = as.Date("1961-01-01"), 
                to = as.Date("1999-12-31"), by = "day")
# Keep non-zero rainfall, exclude 1999 observations
nzrain <- maiquetia[year(day) < 1999 & maiquetia > 0]
```

We will ignore temporal dependence and stationarity, but these should be considered.
The first step in our analysis is to choose a threshold. We select candidates based on quantiles and test whether they lead to stable inference.

```{r}
ths <- quantile(nzrain, seq(0.9, 0.99, length.out = 15))
# Threshold selection diagnostics
W.diag(xdat = nzrain, model = "nhpp", plots = c("WN","PS"), u = ths)
NC.diag(nzrain, u = ths)
```

Threshold selection diagnostics indicate that the shape is more or less constant over the range of threshold considered. Wadsworth's white noise sequence yields a very large residual around $u=20$, but the Northrop--Coleman score test yields $P$-value path above 0.2 for every threshold under consideration. 

```{r}
# Fit using maximum likelihood estimation
gpdf <- fit.gpd(nzrain, threshold = 20, show = TRUE)
```

The default optimization routine for the generalized Pareto distribution is
Grimshaw's method, which profiles out the likelihood. The method is almost guaranteed to converge. Because of non-regularity, the maximum likelihood estimator for $\xi < -1$ does not solve the score equation and leads to infinite log-likelihood, hence the maximum returned lies on the boundary. The standard errors are based on the inverse observed information matrix and provided only if $\xi>-1/2$. We can verify that our maximum likelihood estimate is indeed a maximum by checking if it solves the score equation  if $\hat{\xi}>-1$.

```{r}
isTRUE(all.equal(
  gpd.score(gpdf$estimate, dat = gpdf$exceedances),
  c(0,0), tolerance = 1e-5))
```

If the sample is small, maximum likelihood estimators are biased for the generalized Pareto distribution (the shape parameter is negatively biased, regardless of the true value for $\xi$). Bias correction methods includes the modified score of Firth, but the default method is the  implicit correction (`subtract`), which solves the 
implicit equation 
\begin{align}
   \bs{\tilde{\theta}}=\hat{\bs{\theta}}-\bs{b}(\tilde{\bs{\theta}}). \label{eq:implbias}
\end{align}
The point estimate $\bs{\tilde{\theta}}$ is obtained numerically as the root of this nonlinear system of 
equations. In the present case, the sample size is large and hence the first-order correction, derived through asymptotic arguments from the generalized Pareto distribution likelihood, is small. Note that the bias correction requires $\xi > -1/3$, since it is based on third-order cumulants of the distribution.

```{r}
gpdbcor <- gpd.bcor(dat = gpdf$exceedances, par = gpdf$estimate)
#print the differences between MLE and bias-corrected estimates
gpdf$estimate - gpdbcor
```

The package includes some default diagnostic plots (probability-probability plots and quantile-quantile plots), which include approximate confidence intervals based on order statistics. We can also get profile likelihood and profile-based confidence intervals for most quantities of interest (parameters of the generalized Pareto distribution, excepted shortfall, return levels, $N$-observation maxima mean and quantiles). The example below gives the estimated profile for the median of the centenial maximum distribution conditional on exceeding 15, along with 95% confidence intervals.

```{r}
plot(gpdf, which = 2) #Q-Q plot
# Profile of median of maxima of 100 years
profile <- gpd.pll(param = "Nquant", dat = nzrain, threshold = 15, 
        N = length(nzrain)/(1998-1961)*100, q = 0.5)
# 95% confidence intervals
conf <- confint(profile, print = TRUE)
```
